On the Gravitational stability of gravito-turbulent accretion disks

### 2016-01-25 paper/summary.tex
revising summary

\section{Summary and discussion}\label{summary}
In this paper, we develop the linear theory of cooling, irradiated,
and turbulent accretion disks in order to understand gravitational
instability (GI) in realistic protoplanetary disks (PPDs). 
We model turbulence as an effective, 
Navier-Stokes viscosity. This viscosity provides a 
background heating to balance the imposed cooling,  
and may also be active in the perturbed state.  
We suggest that disk fragmentation, observed in numerical simulations,
can be understood as the non-linear outcome of a secondary
GI of the gravito-turbulent basic state, driven by cooling and/or viscosity.   
%This %allows us to assess realistic disk fragmentation without using
%uncertain empirical conditions based on numerical simulations.   
%{\bf cooling removes pressure support, viscosity removes rotational support}
%{\bf our conclusions do require input of emperical resuls from numerical
%sims}
%going beyond toomre 
%move to subsection?
While viscosity and cooling are related through thermal
balance, they independently enhance GI: cooling reduces thermal
stabilization; and viscous forces compromise rotational
stabilization. This provides a physical explanation as to why
rapidly-cooled, gravito-turbulent disks cannot exist.

@@

The effect of cooling and irradiation on GI is quantified by the
dispersion relation 
Eq. \ref{inviscid}. The sufficient condition for stability 
depends on the irradiation level (Eq. \ref{stable_condition}), and 
\emph{independent} of the cooling rate, provided it is non-zero and
finite. This means that long cooling times can still  
formally lead to instability. However, growth timescales may be 
uninterestingly long for cooling times $\tcool\Omega\gtrsim
O(10)$. Because cooling affects pressure support, GI driven by cooling
occur on small scales, $kH\gtrsim O(1)$.

@@

We generalize the `viscous gravitational instability', previously 
studied in isothermal disks
\citep{lynden-bell74,willerding92,gammie96}, to include 
cooling, viscous heating and irradiation. In our fiducial disk,  
where viscosity and self-gravity are inversely
related ($\alpha\propto Q^{-2}$) to model a
gravito-turbulent background, viscous GI occurs on orbital  
timescales for $\alpha\gtrsim 0.1$. This is consistent with the notion
of a maximum stress sustainable by gravito-turbulence established in
previous numerical simulations \citep{rice05}. Because viscosity affects 
rotational support, viscous GI occurs on large scales, $kH\lesssim
O(1)$.  Furthermore, irradiation preferentially
stabilizes small-scale perturbations. %maybe that's why irradiated disks produce smaller alpha

@@

We apply our linear framework to protoplanetary disks with 
realistic models for cooling and gravito-turbulence
\citep{rafikov15}. We show that, for linear dynamics, realistic 
cooling in a physically non-irradiated disk ($T_\mathrm{irr}=0$) is
actually modeled by `beta' cooling with a  
non-zero irradiation parameter ($\theta\neq0$), in contrast to 
numerical simulations which typically employ beta cooling with  
$\theta=0$ \citep[e.g.][]{gammie01}. How relevant 
are such simulations to realistic PPDs must then be further 
investigated.

The important implication of having 
$\theta\sim 1$ in a PPD is that cooling alone cannot 
destabilize the disk at large radii. Instead, viscous GI occurs on
dynamical timescales for $R\gtrsim 60$AU because 
$\alpha\gtrsim0.1$ there. This corresponds to a Toomre $Q\simeq 1.5$ and 
a cooling time $\tcool\lesssim 3\Omega^{-1}$. These are coincident with 
\emph{empirical} conditions cited in the literature to determine disk
fragmentation 
\citep[e.g.][]{rafikov15}. Here, we attribute a \emph{physical}
cause for the fragmentation of realistic PPDs: gravito-turbulent PPDs
fragment when turbulent stresses are large enough to further
destabilize the disk against self-gravity.    
%collary: details of cooling unimportant as long as the associated
%stress is the same

@@

%we do not look at fragment survival
%fragment due to cooling or viscosity?
%they should produce diff size frags
\subsection{Relation to numerical simulations}\label{prev_works}
Our framework may shed light on 
%may offer an alternative understanding of 
some recent numerical results concerning disk
fragmentation.  \cite{paardekooper12} found that fragmentation can occur for 
slowly-cooled disks, $\tcool\Omega\gg O(1)$, but that this requires  
simulations to run for significantly longer than dynamical 
timescales. This is consistent with our finding
that for either cooling-driven or viscous GI, there is no critical 
cooling rate (viscosity) beyond (below) which the disk is  
formally stable. Instead, growth rates smoothly decrease with
increasing (decreasing) $\tcool$ ($\alpha$), implying that
instabilities take longer to develop for slowly-cooled disks.        
%hopkins paper

%resolution convergence 
Another issue is convergence with numerical resolution. Notably, 
\cite{baehr15} found that for fixed cooling rates, disks eventually 
fragment with sufficient spatial resolution. %This could be due to 
%small-scale, turbulent angular momentum transport being enabled at
%high resolutions, which aids clump formation. 
This might be related to viscous GI in the following sense: high
resolutions enable small-scale turbulent angular momentum transport
away from over-dense regions, aiding clump formation. This 
mechanism is unavailable at low resolutions, for which one can only
resort to rapid cooling for destabilization.

We note that power in gravito-turbulence is dominated by scales with
$kH\sim 1$ \citep{cossins09}, and this is always  
well-resolved by modern numerical simulations. However, this does not
necessarily imply small-scale dynamics are unimportant \citep[especially for non-linear evolution, see][]{young15}, 
as is evident from the non-converging simulations described above.

@@

\subsection{MHD turbulence}%turbulence and viscosity 
It is important to note that while we consider gravito-turbulent disks  
as an application, the linear theory itself does not assume any 
specific origin of turbulence.

>>>>

\section{Summary and discussion}\label{summary}
In this paper, we develop the linear theory of cooling, irradiated,
and viscous accretion disks in order to understand gravitational
instability (GI) in realistic protoplanetary disks (PPDs). 
We use a Navier-Stokes viscosity to mimic the effects of turbulence. 
This viscosity provides a background heating to balance the imposed cooling,  
and may also act on linear perturbations. %thermo and dynamic 
We suggest that disk fragmentation, observed in numerical simulations, 
can be understood as the eventual outcome of secondary
instabilities of the gravito-turbulent base state, driven by cooling
and/or viscosity.     
%This %allows us to assess realistic disk fragmentation without using
%uncertain empirical conditions based on numerical simulations.   
%{\bf cooling removes pressure support, viscosity removes rotational support}
%{\bf our conclusions do require input of emperical resuls from numerical
%sims}
%going beyond toomre 
%move to subsection?
While viscosity and cooling can be related through thermal
balance, they independently enhance growth rates: cooling reduces
thermal stabilization; and viscous forces compromise rotational
stabilization. This provides a physical explanation as to why
rapidly-cooled, gravito-turbulent disks cannot exist.

@@

The effect of cooling and irradiation on GI is quantified by the 
dispersion relation 
Eq. \ref{inviscid}. The sufficient condition for stability 
depends on the irradiation level (Eq. \ref{stable_condition}), and is 
\emph{independent} of the cooling time, provided it is non-zero and
finite. This means that long cooling times can still  
formally lead to instability. However, growth timescales may be 
uninterestingly long for cooling times $\tcool\Omega\gtrsim 
O(10)$. Because cooling affects pressure support, GI driven by cooling
occur on small scales, $kH\gtrsim O(1)$.

@@

We generalize the `viscous gravitational instability', previously 
studied in isothermal disks
\citep{lynden-bell74,willerding92,gammie96}, to include 
cooling, viscous heating and irradiation. In our fiducial disk,  
where viscosity and self-gravity are inversely
related ($\alpha\propto Q^{-2}$) to model a
gravito-turbulent background, this instability occurs on orbital  
timescales for $\alpha\gtrsim 0.1$. This is consistent with the notion
of a maximum stress sustainable by gravito-turbulence established
through numerical simulations \citep{rice05}. Because viscosity affects 
rotational support, viscous GI occurs on large scales, $kH\lesssim
O(1)$.  Furthermore, irradiation preferentially
stabilizes small-scale perturbations. %also true for inviscid case
%maybe that's why irradiated disks produce smaller alpha

@@

We apply our linear framework to protoplanetary disks with 
realistic models for cooling and gravito-turbulence. 
%\citep{rafikov15}. 
%We show that, for linear dynamics, realistic 
%cooling in a physically non-irradiated disk ($T_\mathrm{irr}=0$) is
%actually modeled by `beta' cooling with a  
%non-zero irradiation parameter ($\theta\neq0$), in contrast to 
%numerical simulations which typically employ beta cooling with  
%$\theta=0$ \citep[e.g.][]{gammie01}. How relevant 
%are such simulations to realistic PPDs must then be further 
%investigated. 
%The important implication of having 
%$\theta\sim 1$ in a PPD is that cooling alone cannot 
%destabilize the disk at large radii. 
We show that with a realistic cooling model for PPDs,
cooling alone cannot lead to gravitational instabilities. This is due
to stabilization by an effective irradiation associated with the PPD
cooling function, which is present even if 
there is no physical irradiation. %due to matching functional forms of
                                %cooling 
Instead, viscous GI occurs on  
dynamical timescales for $R\gtrsim 60$AU because 
$\alpha\gtrsim0.1$ there. This corresponds to a Toomre $Q\simeq 1.5$ and 
a cooling time $\tcool\lesssim 3\Omega^{-1}$. These are coincident with 
\emph{empirical} conditions cited in the literature to determine disk
fragmentation 
\citep[e.g.][]{rafikov15}. Here, we attribute a \emph{physical}
cause for the fragmentation of realistic PPDs: gravito-turbulent PPDs
fragment when turbulent stresses are large enough to further
destabilize the disk against self-gravity.    
%collary: details of cooling unimportant as long as the associated
%stress is the same

@@

%we do not look at fragment survival
%fragment due to cooling or viscosity?
%they should produce diff size frags
\subsection{Relation to numerical simulations}\label{prev_works}
%frag in sims because of : cooling, viscosity, or high den
%flucuations? don't know 
Our results may shed light on some numerical simulations 
of disk fragmentation. Table 
\ref{bstar_compare} shows a close match between the characteristic
cooling time for cooling-driven GI (Eq. \ref{betastar}) 
and that for disk fragmentation observed in simulations 
\citep{gammie01,rice05,rice11}. This suggests that, at least for those
simulations, fragmentation is physically due to the removal of
pressure stabilization by cooling on a radial scale of the disk scale
height. 
%i.e. cool-driven g 
{\bf don't know if we want to dwell on this table}

More recent simulations have raised the issue of numerical
convergence. \cite{paardekooper12} found that fragmentation can occur
for   
slowly-cooled disks, $\tcool\Omega\gg O(1)$, but that this requires   
simulations to run for significantly longer than dynamical 
timescales. This is consistent with our finding
that for either cooling-driven or viscous GI, there is no critical 
cooling rate/viscosity below  which the disk is  
formally stable. Instead, growth rates smoothly decrease with
increasing $\tcool$ (decreasing $\alpha$), implying that
instabilities, and hence fragmentation, simply take longer to develop
for slowly-cooled disks.         
%hopkins paper

%resolution convergence 
Another issue is convergence with numerical resolution. For example,
\cite{meru11} found that the critical cooling time for 
disk fragmentation increases with numerical resolution. %sph particles in
% their case 
Follow-up studies attributed this to decreasing numerical viscous
heating at higher resolution, which helps fragmentation
\citep{lodato11,meru12}. %also global effects in
                         %these. (paardekooper/meru/baruteau) 
%note that these sims do not have a floor temp
%is there an effective floor temp related to num visc?
%high res -> smaller effective floor temp -> easier to frag?

Here, we highlight that most numerical experiments, including those
above, employ the standard beta cooling function, Eq. \ref{beta_cool},
\emph{without} a floor temperature. We show in \S\ref{2d_inviscid}
(see also \S\ref{cool_gi}) that this implies cooling-driven GI can
occur at any sufficiently small scales. Then as the numerical
resolution increases, simulations can access a wider range of unstable
scales.  %no upper bound 
In this respect, it is not surprising to find non-convergence with 
increasing resolution. %simple beta cooling will not give a
                       %well-imposed problem

We thus suggest that having a floor temperature in the standard beta
cooling prescription is a necessary condition for  
convergence with respect to numerical resolution. This limits the
relevant scales to a finite range.  %for cool gi 
%physically irradiation 
Furthermore, without a floor temperature, standard beta cooling only
depends on the pressure/energy density field. It is not clear how
results obtained from such models can be applied to PPDs, where
cooling depends on two thermodynamic states (e.g. $P$ and
$T$). A floor temperature permits a mapping between beta
cooling and realistic cooling (\S\ref{ppd_cooling}).

%ultra-small scale cooling-GI may have slow growth rates
%but will eventually matter if you sim for a long time 
%
%limit the relevant scales to a finite range 
%we suggest that having a floor temperature (irradiation) is necessary
%(though may not sufficient) 
%floor temp may be physical: irradiation, or numerical (sph particles
%can't have zero jitter)

%point out viscosity can have a dynamic role in helping frag

However, \cite{baehr15} included a floor temperature (or irradiation)
in their simulations but still find that at fixed cooling rates, disks
eventually fragment with sufficient spatial resolution.
In light of our results on viscous GI, 
we here highlight a possible contribution: %in addition to num visc (meru lodato)
high resolutions enable small-scale turbulent angular momentum transport 
% physical, cf numerical  
to aid clump formation (see also \S\ref{MHD} below). 
This mechanism is unavailable at low resolutions, for which one must
rely on rapid cooling.

%This, in fact, has been suggested for

We note that, although power in gravito-turbulence is dominated by scales with
$kH\sim 1$ \citep{cossins09}, which is  
well-resolved by modern numerical simulations, that this does not
necessarily imply small-scale dynamics/thermodynamics are unimportant
\citep[especially for non-linear evolution, see][]{young15},  
as is evident from the non-converging simulations described above.

@@

\subsection{MHD turbulence}\label{MHD}%turbulence and viscosity 
We emphasize the linear theory we have developed does not assume

%It is important to note that while we consider gravito-turbulent disks  
%as an application, the linear theory itself does not assume any 
%specific origin of turbulence.
